Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-12-22 paper/ms.tex
Revised notation; Lorentzian -> damped sinusoid.

We present a scalable method for Gaussian Process regression in one dimension
with a specific emphasis on large astronomical time-series data sets.
This method can be applied to any Gaussian Process model where the spectral
density can be expressed as any general mixture of Lorentzian functions.

@@

In this paper, we present a class of GP models that enable likelihood
calculations that scale linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must be written as a mixture of Lorentzian functions in
Fourier space.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, robust, simple,
and fast.

@@

In the case of $p$ complex Lorentzians we can check for negative values of the PSD by solving for the
roots of the power spectrum, abbreviating with $z = \omega^2$:
\begin{equation}
P(\omega)=  \sum_{j=1}^{p} \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
\begin{equation}
P_0(z) = \sum_{j=1}^{p} (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
\end{equation}
which is a polynomial with order $p_{ord}=2(p-1)+1$.  With $p=2$, this yields a cubic equation
which may be solved exactly for the roots.

@@

The kernel is modeled as a sum of damped harmonics,
\begin{equation}
k(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)},
\end{equation}
where $\tau > 0$ is a time-offset between data points, $\beta_R$ represents the decay
in correlation with time and $\beta_I$ represents oscillations of period $P = 2\pi/\beta_I$, and
$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
of the measured times if the correlated components are zero.

Now, the damped harmonic kernel components may be rewritten as:
\begin{equation}
 a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
\end{equation}
where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran:2015}.
The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
but otherwise the mathematical form of the extended matrix is identical, but with twice as many
exponential kernel components as the real case.

Although this is a straightforward extension to the \texttt{GenRP} formalism, there is a disadvantage
in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
arithmetic.  In addition, it turns out that the complex conjugate component, $\beta^* = \beta_R - i\beta_I$, in
the extended matrix gives a solution which, not surprisingly, is the complex conjugate of the
component $\beta = \beta_R + i \beta_I$;  thus the computation is redundant.  Finally, complex
arithmetic has the disadvantage that it is not included in automatic-differentiation packages,
which is what we would like to use to compute the derivative of the likelihood (as it is very difficult
to compute this explicitly!).

Consequently, we use the fact that $\exp{(i\beta_I \tau)} =  \cos{(\beta_I \tau)} + i \sin{(\beta_I \tau)}$ to
rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
may be replaced by a $2\times 2$ anti-symmetric matrix,
\begin{equation}
\left(\begin{tabular}{ll}
0 & 1 \\
-1 & 0
\end{tabular}\right),
\end{equation}
while real numbers use the $2\times 2$ identity matrix, $\mathbf{I}$,
so that $\cos{(\beta_I \tau)} + i \sin{(\beta_I \tau)}$ may be rewritten as the rotation matrix
\begin{equation}
\left(\begin{tabular}{ll}
$\cos{(\beta_I \tau)}$ & $\sin{(\beta_I \tau)}$ \\
$-\sin{(\beta_I \tau)}$ & $\cos{(\beta_I \tau)}$
\end{tabular}\right).
\end{equation}
The algebraic properties of these matrices are identical to the complex numbers, but only involve
real components.
We accomplish this transformation in Appendix \ref{appendixa} by taking sums and differences of the complex
equations, resulting in a real extended matrix that is the same size as the complex matrix.
This transformation leads to about a factor of two in computational savings, but has the
additional advantage of allowing the automatic differentiation due to the use of real arithmetic.

@@

The functional form of $k_j$ may be used to approximate some commonly used Kernel components.
The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be
used to approximate each of these kernels to better than 1\% with various numbers of damped
cosine kernel components:
\begin{itemize}
\item {\bf Constant kernel.} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented exactly with $\beta_R = \beta_I =  0$.
\item {\bf Exponential kernel.} This kernel, $k(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
represented with $\beta_I=0$, $\beta_R = \tau_0^{-1}$.
\item {\bf Cosine kernel.} The cosine kernel, $k(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$,
may be represented exactly with $\beta_R=0$ and $\beta_I = \frac{2\pi}{P}$.
\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
Likewise, the Fourier transform of the squared exponential kernel is $\hat G(\omega) = e^{-\omega^2/2}$,
which also declines steeply at large $\omega$;  this doesn't match the more gradual decline of a
Lorentzian.  Nevertheless, the squared exponential kernel may be well approximated by the sum of
four exponential kernels:
\begin{eqnarray}
G(z) &\approx& 63.119512 e^{-1.479579 z} -32.173615 e^{-1.593691 z} -29.425156 e^{-1.388929 z}\\
&+& e^{-1.383928 z} \left[-0.519196\cos{(1.883396 z)} +0.285478 \sin{(1.883396 z)}\right] \\
\end{eqnarray}
which has an accuracy of better than $<3\times 10^{-3}$ for all z, and the power spectrum has
been constrained to be positive.
\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
\begin{equation}
C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
\end{equation}
where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
values of $p$, the Mat\'ern kernel may be approximated well by the sum of
$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel,
$C_{1/2}(x) = \sigma^2 e^{-x}$ ($p=0$), while
$p=1$ is relevant to stellar variability, which we describe next.
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ ($p=1$) kernel is given by:
\begin{equation}
k_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
\end{equation}
where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier transform\footnote{Using the normalization convention:
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat k_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels:
\begin{equation}
k_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
\end{equation}
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error of the approximation scales as $2/(e^2b) \approx 0.27/b$; for
$b=100$, the error in this approximation is $<0.3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

@@

A standard practice in constructing kernels for solving problems with Gaussian processes
is to add or multiply the preceding standard kernel components.  In the case of damped cosine
kernels, the product of two kernels is:
\begin{eqnarray}
k_j(\tau)k_k(\tau) &=&
\frac{1}{2} a_k a_j e^{-(\beta_{R,j}+\beta_{R,k}) \tau} \left(\cos{(\beta_{I,j}-\beta_{I,k})\tau}+\cos{(\beta_{I,j}+\beta_{I,k})\tau}\right)\\
&=& \frac{1}{2} \left(k_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}-\beta_{I,k})\right.\\
&+&\left. k_j(\tau,a_ja_k,\beta_{R,j}+\beta_{R,k},\beta_{I,j}+\beta_{I,k})\right).
\end{eqnarray}
which is the sum of two damped cosine kernels.  Thus, only addition is necessary to
construct new kernels from damped cosine kernels.

The upshot is that the damped cosine is an expressive basis kernel that may be used
to represent a wide range of autocorrelation functions relevant for stellar variability.
We recommend using a series of sums of damped cosines (in which some parameters may be
fixed to specified values to represent particular kernels) to model different amplitudes
and timescales of variability.  If an initial estimate of the power spectrum or
autocorrelation function is available, then these may be fit with this series to
initialize the coefficients of the damped cosine kernel components.
Since a Gaussian Process
must be positive definite, we recommend modeling the logarithm of $a_j$,
$\beta_{R,j}$, and $\beta_{I,j}$ to enforce this to be the case.  One caveat is
that to produce certain kernels, a negative value of $a_j$ is needed (such
as the Mat\'ern 3/2 kernel).  In this case, the value of the sum of
two coefficients may be sampled from a logarithm, while the individual
values are allowed to be negative;  this will enforce the positive definite
constraint, but may allow for other constraints, such as an autocorrelation
function which has a derivative of zero at $\tau = 0$.

@@

@@

\section{Prediction/interpolation}

Precition/interpolation involves constructing a matrix of covariances
between the (noisy) training set and the times at which we wish to
predict or interpolate data points.   If our training times are given by
${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
mean values at the predicted times are given by (RW 2.23):
\begin{eqnarray}
\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
y_k &=& \sum_j k(t_j,t_k) b_j,
\end{eqnarray}
where the latter equation may be solved for ${\bf b}$ using the extended
matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
of $MN$ multiplications is required to obtain the predicted values.
If $M$ is large, this can become quickly prohibitive.  It turns out
that the structure of the sum of exponential kernels may be exploited
to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
i.e.\ a linear scaling with the total number of data points, which can
be much fewer if $M$ is large.

The procedure works as follows:
\begin{itemize}
\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
compute:
\begin{eqnarray} \label{prediction}
\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} b_j\\
\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} b_j,
\end{eqnarray}
where we have separated out the training times before and after the predicted time $t_{*,k}$.

\item Next, we update the coefficients recursively:
\begin{eqnarray}
\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} b_j\right]\\
\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} b_j\right].
\end{eqnarray}
With these updates, we can then use equation \ref{prediction} to compute the expected
mean at $t_{*,k}$ with an additional $p$ operations.  Since there are $N$ values of
$t_k$, the number of operations needed to update $\epsilon$ is of order $p(N+M)$, and
so this procedure avoids a heavy computational burden.

The variance may be handled in a similar manner, but with $p^2$ coefficients
to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
\begin{eqnarray}
cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
\end{eqnarray}
The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
\end{itemize}

Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
just solve for the difference in fewer operations? {\bf TBD}

@@

Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
follow closely the notation and equations given in \citet{Ambikasaran:2015}.  We start
with the observation that a Lorentzian correlation function may be decomposed into the sum of two
exponential functions:
\begin{eqnarray}
k(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
 &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
\end{eqnarray}
where $\beta = \beta_R + i\beta_I$ is the
complex kernel parameter (i.e.\ $\beta_R = Re(\beta)$, $\beta_I = Im(\beta)$), $\omega = -\beta_I = 2\pi/P$
is the frequency of oscillation of the kernel with period $P$, and $x^*$ is the complex conjugate of $x$.

Now, the Lorentzian kernal may be included in the GRP algorithm
using the two exponential kernels with $\beta_1 = \beta$ and $\beta_2 = \beta^*$.  This
requires the algorithm to use complex arithmetic, which has several disadvantages:
1).\ complex arithmetic is more expensive by a factor of 2-3; 2).\ the complex conjugate equation
is redundant as it turns out that $l_2 = l_1^*$, $r_2 = r_1^*$, and $x_i = x_i^*$ (i.e.\ $x_i$ is
real), meaning that the same equations are being solved twice for the real and complex components;
and 3).\ if the derivative of the likelihood is computed with automatic
differentiation (AD), most AD algorithms require real variables.

>>>>

We present a scalable method for Gaussian Process regression in one dimension
with a specific emphasis on large astronomical time-series data sets.
This method can be applied to any Gaussian Process model where the spectral
density can be expressed as any general mixture of damped sinusoid functions.

@@

In this paper, we present a class of GP models that enable likelihood
calculations that scale linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must be written as a mixture of damped sinusoid functions.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, robust, simple,
and fast.

@@

In the case of $p$ damped sinusoids we can check for negative values of the PSD by solving for the
roots of the power spectrum, abbreviating with $z = \omega^2$:
\begin{equation}
P(\omega)=  \sum_{j=1}^{p} \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
\begin{equation}
P_0(z) = \sum_{j=1}^{p} (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
\end{equation}
which is a polynomial with order $p_{ord}=2(p-1)+1$.  With $p=2$, this yields a cubic equation
which may be solved exactly for the roots.

@@

%The kernel is modeled as a sum of damped harmonics,
%\begin{equation}
%k_j(\tau) = a_0 \delta(\tau) + \sum_{j=1}^{N_{kernel}} a_j \exp{(-b_j \tau)} \cos{(-d_j \tau)} + b_j \exp{(-b_j \tau)} \sin{(-d_j \tau)},
%\end{equation}
%where $\tau > 0$ is a time-offset between data points, $c_j$ represents the decay
%in correlation with time and $d_j$ represents oscillations of period $P = 2\pi/d_j$, and
%$a_0 = \sigma^2$ is a white-noise component with $\delta(0) = 1$ and zero otherwise.  Formally
%the $\delta(\tau)$ function is a Kronecker delta function for which $\tau$ is the midpoint in
%time measured over a finite duration exposure time, and $\sigma$ is the standard deviation
%of the measured times if the correlated components are zero.

%Now, the damped harmonic kernel components may be rewritten as:
%\begin{equation}
% a_j e^{-\beta_R \tau} \cos{(\beta_I \tau)} = a_j \frac{1}{2}\left( e^{-(\beta_R + i \beta_I) \tau} +  e^{-(\beta_R + i \beta_I) \tau}\right),
%\end{equation}
%where $i^2 = -1$, which has the advantage of utilizing precisely the same form as the Rybicki-Press covariance
%matrix, and thus the Generalized Rybicki-Press algorithm may be applied as in \citet{Ambikasaran:2015}.
%The only difference is that the $\beta$ coefficients are now complex, $\beta = \beta_R + i \beta_I$,
%but otherwise the mathematical form of the extended matrix is identical, but with twice as many
%exponential kernel components as the real case.

Although the complex kernel is a straightforward extension to the \texttt{GenRP} formalism, there is a disadvantage
in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
arithmetic.  In addition, it turns out that the complex conjugate component, $(c_j+id_j)^* = c_j - id_j$, in
the extended matrix gives a solution which, not surprisingly, is the complex conjugate of the
component $c_j + i d_j$;  thus the computation is redundant.  Finally, complex
arithmetic has the disadvantage that it is not included in automatic-differentiation packages,
which is what we would like to use to compute the derivative of the likelihood (as it is very difficult
to compute this explicitly!).

Consequently, we use the fact that $\exp{(id_j\tau)} =  \cos{(d_j\tau)} + i \sin{(d_j \tau)}$ to
rewrite the extended matrix in terms of real arithmetic using the fact that the imaginary number $i$
may be replaced by a $2\times 2$ anti-symmetric matrix,
\begin{equation}
\left(\begin{tabular}{ll}
0 & 1 \\
-1 & 0
\end{tabular}\right),
\end{equation}
while real numbers use the $2\times 2$ identity matrix, $\mathbf{I}$,
so that $\cos{(d_j\tau)} + i \sin{(d_j\tau)}$ may be rewritten as the rotation matrix
\begin{equation}
\left(\begin{tabular}{ll}
$\cos{(d_j\tau)}$ & $\sin{(d_j\tau)}$ \\
$-\sin{(d_j\tau)}$ & $\cos{(d_j \tau)}$
\end{tabular}\right).
\end{equation}
The algebraic properties of these matrices are identical to the complex numbers, but only involve
real components.
We accomplish this transformation in Appendix \ref{appendixa} by taking sums and differences of the complex
equations, resulting in a real extended matrix that is the same size as the complex matrix.
This transformation leads to about a factor of two in computational savings, but has the
additional advantage of allowing the automatic differentiation due to the use of real arithmetic.

@@

The functional form of $k_j$ may be used to approximate some commonly used Kernel components.
The squared exponential (Gaussian) kernel and the Mat\'ern kernels are commonly
used radial kernels, while the cosine kernel and the exponential-sine-squared kernels are
commonly used periodic kernels.  Linear combinations of our basis of damped cosine kernels may be
used to approximate each of these kernels to better than 1\% with various numbers of damped
cosine kernel components:
\begin{itemize}
\item {\bf Constant kernel.} A constant kernel component can be used to capture long-timescale
correlations in data, and can be represented exactly with $c_j= d_j =  0$.
\item {\bf Exponential kernel.} This kernel, $k(\tau) = e^{-\vert \tau/\tau_0\vert}$, is exactly
represented with $d_j=0$, $c_j= \tau_0^{-1}$.
\item {\bf Cosine kernel.} The cosine kernel, $k(t) = \cos{\left(\frac{2\pi}{P} \vert \tau \vert\right)}$,
may be represented exactly with $c_j=0$ and $d_j= \frac{2\pi}{P}$.
\item {\bf Squared exponential.} The squared exponential or Gaussian kernel, $G(z)=e^{-z^2/2}$,
has the behavior of declining steeply at large time separation, is an even function, and has $c_1$ = 0.
Likewise, the Fourier transform of the squared exponential kernel is $\hat G(\omega) = e^{-\omega^2/2}$,
which also declines steeply at large $\omega$;  this doesn't match the more gradual decline of the
power spectrum of a damped sinuosoid.  Nevertheless, the squared exponential kernel may be well approximated 
by the sum of four exponential kernels:
\begin{eqnarray}
G(z) &\approx& 63.119512 e^{-1.479579 z} -32.173615 e^{-1.593691 z} -29.425156 e^{-1.388929 z}\\
&+& e^{-1.383928 z} \left[-0.519196\cos{(1.883396 z)} +0.285478 \sin{(1.883396 z)}\right] \\
\end{eqnarray}
which has an accuracy of better than $<3\times 10^{-3}$ for all z, and the power spectrum has
been constrained to be positive.
\item {\bf Mat\'ern $p+1/2$ kernel.}  The Mat\'ern kernel is given by:
\begin{equation}
C_{p+1/2}(x) = \sigma^2 e^{-\sqrt{(2p+1)}x} \frac{\Gamma(p+1)}{\Gamma(2p+1)}
\sum_{i=0}^p \frac{(p+i)!}{i!(p-i)!} (\sqrt{(8p+4)}x)^{p-i},
\end{equation}
where $p$ is a positive integer.  In the limit $p\rightarrow \infty$, this becomes the
squared exponential kernel: $C_\infty(x) = \sigma^2 e^{-x^2/2}$.  But, for small
values of $p$, the Mat\'ern kernel may be approximated well by the sum of
$2p$ exponential kernels.  The exponential kernel is the Mat\'ern 1/2 kernel,
$C_{1/2}(x) = \sigma^2 e^{-x}$ ($p=0$), while
$p=1$ is relevant to stellar variability, which we describe next.
\item {\bf Mat\'ern 3/2 kernel.}  The Mat\'ern $3/2$ ($p=1$) kernel is given by:
\begin{equation}
k_{3/2}(\tau) = C_{3/2}(\tau/\tau_0) = (1+v)e^{-v},
\end{equation}
where $v = \sqrt{3}\vert \tau/\tau_0\vert$.  This kernel has a Fourier transform\footnote{Using the normalization convention:
$\hat f(\omega) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^\infty f(t) e^{i\omega t} dt$}  of
\begin{equation}
\hat k_{3/2}(\omega) = \frac{\tau_0 6^{3/2}}{\pi^{1/2}} (3+\tau_0^2\omega^2)^{-2},
\end{equation}
where $\omega = 2\pi f$.  This kernel may be well approximated by the sum of two exponential kernels:
\begin{equation}
k_{3/2}(\tau) \approx (1-b)e^{-v}+be^{-v\frac{b-1}{b}},
\end{equation}
where $b \gg 1$ is a dimensionless parameter; this approximation is exact in the limit $b \rightarrow \infty$.  For
large values of $b$, the maximum error of the approximation scales as $2/(e^2b) \approx 0.27/b$; for
$b=100$, the error in this approximation is $<0.3$\%.  Note that the two exponential terms
have opposite signs, which may lead to numerical instability for values of $b$ that
are large.

@@

A standard practice in constructing kernels for solving problems with Gaussian processes
is to add or multiply the preceding standard kernel components.  In the case of damped cosine
kernels, the product of two kernels is:
\begin{eqnarray}
k_j(\tau)k_k(\tau) &=&
\frac{1}{2} e^{-(c_j+c_k) \tau} \left((a_ja_k+b_jb_k)\cos{\left[(d_j-d_k)\tau\right]} + (a_jb_k-a_kb_j)\sin{\left[(d_j-d_k)\tau\right]} \right. \\
 &+&\left.(a_ja_k-b_jb_k)\cos{\left[(d_j+d_k)\tau\right]} -(a_jb_k-a_kb_j)\sin{\left[(d_j+d_k)\tau\right]}\right),
\end{eqnarray}
which is the sum of two damped sinusoidal kernels.  Thus, only addition is necessary to
construct new kernels from the product of damped sinuosidal kernels.

The upshot is that the damped sinusoid is an expressive basis kernel that may be used
to represent a wide range of autocorrelation functions relevant for stellar variability, and
more.
We recommend using a series of sums of damped sinusoids (in which some parameters may be
fixed to specified values to represent particular kernels) to model different amplitudes
and timescales of variability.  If an initial estimate of the power spectrum or
autocorrelation function is available, then these may be fit with this series to
initialize the coefficients of the damped sinusoid kernel components.
Since a Gaussian Process
must be positive definite, we recommend modeling the logarithm of $a_j$, $b_j$,
$c_j$, and $d_j$ to enforce this to be the case.  One caveat is
that to produce certain kernels, a negative value of $a_j$ or $b_j$ is needed (such
as the Mat\'ern 3/2 kernel).  In this case, the value of the sum of
two coefficients may be sampled from a logarithm, while the individual
values are allowed to be negative;  this will enforce the positive definite
constraint, but may allow for other constraints, such as an autocorrelation
function which has a derivative of zero at $\tau = 0$.

@@

Outside of astronomy, this technique may have application to seismology \citep{Robinson1967},

@@

%\section{Prediction/interpolation}
%
%Precition/interpolation involves constructing a matrix of covariances
%between the (noisy) training set and the times at which we wish to
%predict or interpolate data points.   If our training times are given by
%${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
%mean values at the predicted times are given by (RW 2.23):
%\begin{eqnarray}
%\bar f_{*,k} &=& \sum_j k(t_{*,k},t_j) b_j,\\
%&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
%&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
%y_k &=& \sum_j k(t_j,t_k) h_j,
%\end{eqnarray}
%where the latter equation may be solved for ${\bf h}$ using the extended
%matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
%matrix $k({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
%of $MN$ multiplications is required to obtain the predicted values.
%If $M$ is large, this can become quickly prohibitive.  It turns out
%that the structure of the sum of exponential kernels may be exploited
%to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
%i.e.\ a linear scaling with the total number of data points, which can
%be much fewer if $M$ is large.
%
%The procedure works as follows:
%\begin{itemize}
%\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
%\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
%compute:
%\begin{eqnarray} \label{prediction}
%\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
%\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} h_j\\
%\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} h_j,
%\end{eqnarray}
%where we have separated out the training times before and after the predicted time $t_{*,k}$.
%
%\item Next, we update the coefficients recursively:
%\begin{eqnarray}
%\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} h_j\right]\\
%\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} h_j\right].
%\end{eqnarray}
%With these updates, we can then use equation \ref{prediction} to compute the expected
%mean at $t_{*,k}$ with an additional $p$ operations.  Since there are $N$ values of
%$t_k$, the number of operations needed to update $\epsilon$ is of order $p(N+M)$, and
%so this procedure avoids a heavy computational burden.
%
%The variance may be handled in a similar manner, but with $p^2$ coefficients
%to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
%\begin{eqnarray}
%cov(f_{*,k}) &=&  k(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
%&=& k(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
%\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} k({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
%&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
%\end{eqnarray}
%The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
%\end{itemize}
%
%Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
%then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
%just solve for the difference in fewer operations? {\bf TBD}

@@

Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
follow closely the notation and equations given in \citet{Ambikasaran:2015}.  We start
with the observation that a damped sinusoid correlation function may be decomposed into the sum of two
exponential functions:
\begin{eqnarray}
k_j(\tau) &=& a_j e^{-c_j\tau} \cos{(-d_j\tau)} + b_j e^{-c_j\tau} \sin{(-d_j \tau)}\\
 &=& \frac{1}{2}(a_j+ib_j)\left[e^{-(c_j+id_j) \tau}+e^{-(c_j-id_j) \tau}\right],
\end{eqnarray}
where $c_j + id_j$ is the complex kernel parameter and $d_j = 2\pi/P$
is the frequency of oscillation of the kernel with period $P$.

Now, the damped sinusoid kernal may be included in the GRP algorithm
using the above two exponential kernels.  This
requires the algorithm to use complex arithmetic, which has several disadvantages:
1).\ complex arithmetic is more expensive by a factor of 2-3; 2).\ the complex conjugate equation
is redundant as it turns out that $l_2 = l_1^*$, $r_2 = r_1^*$, and $x_i = x_i^*$ (i.e.\ $x_i$ is
real), meaning that the same equations are being solved twice for the real and complex components;
and 3).\ if the derivative of the likelihood is computed with automatic
differentiation (AD), most AD algorithms require real variables.
